{Gravitational wave sources in our Galactic backyard:}\\{Predictions for BHBH, BHNS and NSNS binaries detectable with LISA

### 2022-03-17 paper/tex/sections/comparisons.tex
further updates after review

\paragraph{\citet{Lau+2020}} focussed on the number of Galactic NSNS binaries that could be detected by LISA. Their study uses the same population synthesis code, COMPAS, as this work, though an earlier version. Despite this, their study finds a much larger number of detections. They make several different physical assumptions in their population synthesis, using the \citet{Fryer+2012} \textit{rapid} remnant mass prescription, assuming the optimistic CE scenario, limiting the maximum neutron star mass to $2 \unit{M_{\odot}}$ and not implementing PISN. However, we note that none of these assumptions strongly affect the NSNS LISA detection rate (see bottom panel of Fig.~\ref{fig:detection_rates}, models \modOpt{}, \modRapid{}, \modNSLow{} and \modNoPISN{}) and so this is unlikely to entirely account for the differences. It is also important to highlight that COMPAS has received several improvements and bug fixes since \citet{Vigna-Gomez+2018} (which contains the simulations used by \citealt{Lau+2020}) and these could possibly have affected the formation rate of NSNSs.
%
Yet it is most likely that the remaining difference between our results is due to way in which we simulate the Milky Way. \citet{Lau+2020} use a model for the Milky Way similar to that of \citet{Breivik+2020}, which we use to estimate the impact of the choice of MW model in Appendix~\ref{app:mw_changes}. The Milky Way model by \cite{Breivik+2020} applies only two metallicity bins, while we consider a range of metallicities between $10^{-4}$ and $0.03$. When applying the simpler model for the Milky Way, we find that the NSNS detection rate is increased by at least a factor of two. \citet{Lau+2020} only uses a single metallicity, and so assuming that all star formation happens at a single high metallicity (which has a high efficiency of producing NSNS), could lead to an even greater overestimate of the detection rate.  Hence, we expect the low number of metallicity bins in their Milky Way model to be the main driver behind the discrepancy between our results.

\paragraph{\citet{Breivik+2020}} introduced the population synthesis code COSMIC and presented detections for many different DCO types in LISA using this code. They find that LISA will detect 93 BHBHs, 33 BHNSs and 8 NSNSs in the Milky Way over a 4 year mission. \citet{Breivik+2020} make many physical assumptions that differ from our fiducial model, the most notable being that they assume the optimistic CE scenario and that case BB mass transfer is always unstable, whilst also using a simpler model for the Milky Way (see Appendix~\ref{app:mw_changes}). Thus for better comparison we ran our simulation using model \modCaseBBOpt{} and the Milky Way model from \citet{Breivik+2020}. This results in 97, 101 and 43 detections for BHBHs, BHNSs and NSNSs respectively. Therefore, though we are in very good agreement for BHBHs, we predict much higher rates for BHNSs and NSNSs. These differences are likely due to using a different population synthesis code (COSMIC), which has different underlying physics assumptions from COMPAS. Given our strong agreement for BHBHs, it is possible that COSMIC and COMPAS handle NSs differently and so lead to different detection rates for DCOs containing NSs. However checking this would require a more in-depth study of the intrinsic formation rate of DCOs containing NSs in the two codes.

>>>>

\paragraph{\citet{Lau+2020}} focussed on the number of Galactic NSNS binaries that could be detected by LISA. Their study uses the same population synthesis code, COMPAS, as this work, though an earlier version. Despite this, their study finds a much larger number of detections. They make several different physical assumptions in their population synthesis, using the \citet{Fryer+2012} \textit{rapid} remnant mass prescription, assuming the optimistic CE scenario, limiting the maximum neutron star mass to $2 \unit{M_{\odot}}$ and not implementing PISN. However, we note that none of these assumptions strongly affect the NSNS LISA detection rate (see bottom panel of Fig.~\ref{fig:detection_rates}, models \modOpt{}, \modRapid{}, \modNSLow{} and \modNoPISN{}) and so this is unlikely to entirely account for the differences. It is also important to highlight that COMPAS has received several improvements and bug fixes since \citet{Vigna-Gomez+2018} (which contains the simulations used by \citealt{Lau+2020}) and these could possibly have affected the formation rate of NSNSs.

Yet it is most likely that the remaining difference between our results is due to way in which we simulate the Milky Way. \citet{Lau+2020} use a model for the Milky Way similar to that of \citet{Breivik+2020}, which we use to estimate the impact of the choice of MW model in Appendix~\ref{app:mw_changes}. The Milky Way model by \cite{Breivik+2020} applies only two metallicity bins, while we consider a range of metallicities between $10^{-4}$ and $0.03$. When applying the simpler model for the Milky Way, we find that the NSNS detection rate is increased by at least a factor of two. \citet{Lau+2020} only uses a single metallicity, and so assuming that all star formation happens at a single high metallicity (which has a high efficiency of producing NSNS), could lead to an even greater overestimate of the detection rate.  Hence, we expect the low number of metallicity bins in their Milky Way model to be the main driver behind the discrepancy between our results.

\paragraph{\citet{Breivik+2020}} introduced the population synthesis code COSMIC and presented detections for many different DCO types in LISA using this code. They find that LISA will detect \edit1{72 BHBHs, 19 BHNSs and 10 NSNSs} in the Milky Way over a 4 year mission. \citet{Breivik+2020} make many physical assumptions that differ from our fiducial model, the most notable being that they assume the optimistic CE scenario and that case BB mass transfer is always unstable, whilst also using a simpler model for the Milky Way (see Appendix~\ref{app:mw_changes}). Thus for better comparison we ran our simulation using model \modCaseBBOpt{} and the Milky Way model from \citet{Breivik+2020}. This results in 97, 101 and 43 detections for BHBHs, BHNSs and NSNSs respectively. \edit1{Therefore, though our BHBH rates agree reasonably well, we predict much higher rates for BHNSs and NSNSs}. These differences are likely due to using a different population synthesis code (COSMIC), which has different underlying physics assumptions from COMPAS. Given our strong agreement for BHBHs, it is possible that COSMIC and COMPAS handle NSs differently and so lead to different detection rates for DCOs containing NSs. However checking this would require a more in-depth study of the intrinsic formation rate of DCOs containing NSs in the two codes.

### 2022-03-17 paper/tex/sections/discussion.tex
further updates after review

\subsection{Identification of GW sources}\label{sec:identify_sources}
It is important to note that, though we present predictions for the detection rates of specific DCO types, the nature of the source may not be immediately \referee{(or ever)} apparent from the gravitational wave signal. LISA can detect a variety of sources, from exoplanets \citep[e.g.][]{Tamanini+2019} to common-envelopes \citep[e.g.][]{Ginat+2020, Renzo+2021} that may cause confusion. However, by far the most prominent will be the population of Galactic WDWDs detectable with LISA, which will be several orders of magnitude larger than the population of the more massive DCOs that we focus on in this paper \citep[e.g.][]{Korol+2017}. It is therefore imperative that we consider how to distinguish NS and BH binaries from this much more numerous population of sources. In addition to distinguishing them from WDWDs, we must consider how to discriminate between BHBHs, BHNSs and NSNSs themselves.

@@

We also note that, for reasons of computational efficiency, we have not accounted for the spatial velocities resulting from the Blaauw-Boersma kick \citep{Blaauw+1961,Boersma1961}. In test simulations we find that accounting for this spreads out the population (increasing the typical height above the Galactic plane and Galactocentric radius), but we find that the impact on the rate is limited. In light of the other much larger uncertainties, we felt that this was justified (see however, e.g., \citealt{Brandt+1995, Abbott+2017_GW170817_progenitor}). We have further ignored a possible contribution coming from the Galactic halo, as \citet{Lamberts+2018} estimates this not be significant. However, this may not be true for other formation channels other than those we have considered here.

>>>>

\subsection{Identification of GW sources}\label{sec:identify_sources}
It is important to note that, though we present predictions for the detection rates of specific DCO types, the nature of the source may not be immediately \edit1{(or ever)} apparent from the gravitational wave signal. LISA can detect a variety of sources, from exoplanets \citep[e.g.][]{Tamanini+2019} to common-envelopes \citep[e.g.][]{Ginat+2020, Renzo+2021} that may cause confusion. However, by far the most prominent will be the population of Galactic WDWDs detectable with LISA, which will be several orders of magnitude larger than the population of the more massive DCOs that we focus on in this paper \citep[e.g.][]{Korol+2017}. It is therefore imperative that we consider how to distinguish NS and BH binaries from this much more numerous population of sources. In addition to distinguishing them from WDWDs, we must consider how to discriminate between BHBHs, BHNSs and NSNSs themselves.

@@

We also note that, for reasons of computational efficiency, we have not accounted for the spatial velocities resulting from the Blaauw-Boersma kick \citep{Blaauw+1961,Boersma1961}. In test simulations we find that accounting for this spreads out the population (increasing the typical height above the Galactic plane and Galactocentric radius), but we find that the impact on the rate is limited. In light of the other much larger uncertainties, we felt that this was justified (see however, e.g., \citealt{Brandt+1995, Abbott+2017_GW170817_progenitor}). \edit1{We have further ignored a possible contribution coming from the Galactic halo, as \citet{Sesana+2020} estimates this is not significant compared to the contributions from the Galactic bulge and discs}. However, this may not be true for other formation channels other than those we have considered here.

### 2022-03-17 paper/tex/sections/introduction.tex
further updates after review

The Laser Interferometer Space Antenna (LISA, \citealp{Amaro-Seoane+2017, Colpi+2019}) will provide observations in an entirely new regime of gravitational waves. LISA will observe at lower frequencies ($10^{-5} \lesssim f / \unit{Hz} \lesssim 10^{-1}$) than ground-based detectors and so will enable the study of sources that are imperceptible by ground-based detectors, such as the mergers of supermassive black holes and extreme mass-ratio inspirals \citep[e.g.][]{Begelman+1980, Klein+2016}. Moreover, this frequency regime is also of interest for the detection of \textit{local} stellar-mass double compact objects (DCOs) millions of years before their merger. This presents an opportunity for both multi-messenger detections to search for electromagnetic counterparts, as well as multiband gravitational-wave detections that can help to constrain binary characteristics \citep[e.g.][]{Sesana+2016, Gerosa+2019}. In addition, LISA will be able to measure the eccentricities of DCOs, which may yield further constraints on binary evolution, differentiate between formation channels and distinguish between DCO types \citep[e.g.][]{Nelemans+2001, Breivik+2016, Antonini+2017, Rodriguez+2018}. Unlike ground-based detectors, \referee{the majority of stellar-mass sources detected in LISA will reside in local galaxies, with most situated in the Milky Way}. These sources could be used as a probe for the structure of our galaxy \citep[e.g.][]{Korol+2019}.

>>>>

The Laser Interferometer Space Antenna (LISA, \citealp{Amaro-Seoane+2017, Colpi+2019}) will provide observations in an entirely new regime of gravitational waves. LISA will observe at lower frequencies ($10^{-5} \lesssim f / \unit{Hz} \lesssim 10^{-1}$) than ground-based detectors and so will enable the study of sources that are imperceptible by ground-based detectors, such as the mergers of supermassive black holes and extreme mass-ratio inspirals \citep[e.g.][]{Begelman+1980, Klein+2016}. Moreover, this frequency regime is also of interest for the detection of \textit{local} stellar-mass double compact objects (DCOs) millions of years before their merger. This presents an opportunity for both multi-messenger detections to search for electromagnetic counterparts, as well as multiband gravitational-wave detections that can help to constrain binary characteristics \citep[e.g.][]{Sesana+2016, Gerosa+2019}. In addition, LISA will be able to measure the eccentricities of DCOs, which may yield further constraints on binary evolution, differentiate between formation channels and distinguish between DCO types \citep[e.g.][]{Nelemans+2001, Breivik+2016, Antonini+2017, Rodriguez+2018}. Unlike ground-based detectors, \edit1{the majority of stellar-mass sources detected in LISA will reside in local galaxies, with most situated in the Milky Way}. These sources could be used as a probe for the structure of our galaxy \citep[e.g.][]{Korol+2019}.

### 2022-03-17 paper/tex/sections/methods.tex
further updates after review

We match each set of galaxy parameters $\mathbf{g}_{{j}}$, to a random set of binary parameters $\mathbf{b}_{{Z, i}}$, by drawing a binary from the closest metallicity bin to the metallicity in $\mathbf{g}_{{j}}$. If the metallicity in $\mathbf{g}_{{j}}$ is below the minimum COMPAS metallicity bin ($Z = 10^{-4}$), we use this minimum bin. If the metallicity in $\mathbf{g}_{{j}}$ is above the maximum COMPAS metallicity bin ($Z = 0.03$), we use a randomly selected bin from the five highest metallicity bins.

@@

We define a binary as detectable if its gravitational wave signal has a signal-to-noise ratio (SNR) of greater than 7 \citep[e.g.][]{Breivik+2020, Korol+2020}. The sky-, polarisation- and orientation-averaged signal-to-noise ratio, $\rho$, of an inspiraling binary can be calculated with the following \citep[e.g.][]{Finn+2000}
\begin{equation}\label{eq:snr}
    \rho^2 = \sum_{n=1}^{\infty} \int_{f_{n, i}}^{f_{n, f}} \frac{h_{c, n}^{2}}{f_{n}^{2} S_{\rm n}\left(f_{n}\right)} \dd{f_n},
\end{equation}
where $n$ is a harmonic of the gravitational wave signal, $f_n = n \cdot f_{\rm orb}$ is the frequency of the $n^{\rm th}$ harmonic of the gravitational wave signal, $f_{\rm orb}$ is the orbital frequency, $S_{\rm n}(f_n)$ is the LISA sensitivity curve at frequency $f_n$ \citep[e.g.][]{Robson+2019} and $h_{c,n}$ is the characteristic strain of the $n^{\rm th}$ harmonic, given by \citep[e.g.][]{Barack+2004}
\begin{equation}\label{eq:charstrain}
    h^2_{c,n} = \frac{2^{5/3}}{3 \pi^{4/3}} \frac{(G \mathcal{M}_c)^{5/3}}{c^3 D_L^2} \frac{1}{f_{\rm orb}^{1/3}} \frac{g(n,e)}{n F(e)},
\end{equation}
where $D_L$ is the luminosity distance to the source, $f_{\rm orb}$ is the orbital frequency, $g(n, e)$ and $F(e)$ are given in \citet{Peters+1963} and $\mathcal{M}_c$ is the chirp mass, defined as
\begin{equation}\label{eq:chirp_mass}
    \mathcal{M}_c = \frac{(m_1 m_2)^{3/5}}{(m_1 + m_2)^{1/5}}.
\end{equation}

>>>>

We match each set of galaxy parameters $\mathbf{g}_{{j}}$, to a random set of binary parameters $\mathbf{b}_{{Z, i}}$, by drawing a binary from the closest metallicity bin to the metallicity in $\mathbf{g}_{{j}}$. If the metallicity in $\mathbf{g}_{{j}}$ is below the minimum COMPAS metallicity bin ($Z = 10^{-4}$), we use this minimum bin. If the metallicity in $\mathbf{g}_{{j}}$ is above the maximum COMPAS metallicity bin ($Z = 0.03$), we use a randomly selected bin from the five highest metallicity bins\footnote{\edit1{We spread the binaries over the five highest bins, rather than just the highest bin, as we found that using a single bin led to unphysical artifacts in our results. These artifacts arose because the small population of binaries in the highest bin were oversampled.}}.

@@

We define a binary as detectable if its gravitational wave signal has a signal-to-noise ratio (SNR) of greater than 7 \edit1{by the end of the LISA mission} \citep[e.g.][]{Breivik+2020, Korol+2020}. The sky-, polarisation- and orientation-averaged signal-to-noise ratio, $\rho$, of an inspiraling binary can be calculated with the following \citep[e.g.][]{Finn+2000}
\begin{equation}\label{eq:snr}
    \rho^2 = \sum_{n=1}^{\infty} \int_{f_{n, i}}^{f_{n, f}} \frac{h_{c, n}^{2}}{f_{n}^{2} S_{\rm n}\left(f_{n}\right)} \dd{f_n},
\end{equation}
where $n$ is a harmonic of the gravitational wave signal, $f_n = n \cdot f_{\rm orb}$ is the frequency of the $n^{\rm th}$ harmonic of the gravitational wave signal, $f_{\rm orb}$ is the orbital frequency, $S_{\rm n}(f_n)$ is the LISA sensitivity curve at frequency $f_n$ \citep[e.g.][]{Robson+2019} and $h_{c,n}$ is the characteristic strain of the $n^{\rm th}$ harmonic, given by \citep[e.g.][]{Barack+2004}
\begin{equation}\label{eq:charstrain}
    h^2_{c,n} = \frac{2^{5/3}}{3 \pi^{4/3}} \frac{(G \mathcal{M}_c)^{5/3}}{c^3 D_L^2} \frac{1}{f_{\rm orb}^{1/3}} \frac{g(n,e)}{n F(e)},
\end{equation}
where $D_L$ is the luminosity distance to the source, $f_{\rm orb}$ is the orbital frequency, $g(n, e)$ and $F(e)$ are given in \citet{Peters+1963} and $\mathcal{M}_c$ is the chirp mass, defined as
\begin{equation}\label{eq:chirp_mass}
    \mathcal{M}_c = \frac{(m_1 m_2)^{3/5}}{(m_1 + m_2)^{1/5}}.
\end{equation}
